---
title: "Data availability"
description: |
  Data availability depends on sources
author:
  - name: Giorgio Comai 
    url: https://giorgiocomai.eu
    affiliation: OBCT/EDJNet
    affiliation_url: https://www.europeandatajournalism.eu/
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library("dplyr", warn.conflicts = FALSE)

library("tidyverse", quietly = TRUE)
library("sf", quietly = TRUE)
library("latlon2map")
library("RSQLite") # for caching
options(timeout = 60000)

cache_folder <- fs::path(fs::path_home_r(),
                         "R",
                         "ll_data")

fs::dir_create(cache_folder)

ll_set_folder(path = cache_folder)

## set db
db <- DBI::dbConnect(
  drv = RSQLite::SQLite(),
  fs::path(cache_folder, "pop_weighted_centre.sqlite")
)

source("functions.R")
```



```{r population grid data availability}
available_pop_grid_years <- c(2006, 2011, 2018)

pop_grid_available <- purrr::map_dfr(.x = available_pop_grid_years,
               .f = function(x) {
  get_pop_grid_countries(x, db) %>% 
                   dplyr::mutate(year = x)
})
pop_grid_available %>% dplyr::group_by(year) %>% count() %>% knitr::kable()

```


## LAU coverage by year


[See source](https://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/population-distribution-demography/geostat) for more details.

https://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units/lau

```{r lau data availability}
available_lau_years <- 2011:2020

lau_available <- purrr::map_dfr(.x = available_lau_years,
               .f = function(x) {
  get_lau_countries(x, db) %>% 
                   dplyr::mutate(year = x)
})
lau_available %>% dplyr::group_by(year) %>% count() %>% knitr::kable()
```


```{r}
lau_available %>%
  dplyr::group_by(year) %>%
  dplyr::arrange(country) %>% 
  summarise(country = stringr::str_c(country, collapse = ", ")) %>% knitr::kable()
```

For LAU, the most recent dataset is not the one with the widest coverage. Largest coverage may be obtained by using older LAU when the most recent are not available for relevant countries. 


```{r}
latest_lau_countries <- lau_available %>% 
  dplyr::slice_max(year) %>% 
  dplyr::pull(country)

all_lau_countries <- lau_available %>% 
#  dplyr::filter(year!= max (year)) %>% 
  dplyr::pull(country) %>% 
  unique()

all_lau_countries[!is.element(all_lau_countries, latest_lau_countries)]
```




## Population grid coverage by year

[See source](https://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/population-distribution-demography/geostat) for more details.

```{r}
pop_grid_available %>%
  dplyr::group_by(year) %>%
  dplyr::arrange(country) %>% 
  summarise(country = stringr::str_c(country, collapse = ", ")) %>%
  knitr::kable()
```




```{r get base data for each source dataset, eval=FALSE}
pop_grid_countries <- get_pop_grid_countries(pop_grid_year, db)
lau_countries <- get_lau_countries(lau_year, db)
gisco_id_df <- get_gisco_id(lau_year, db)
```


## How many countries are present in both the LAU and population grid dataset?

```{r}
pop_grid_lau_countries <- 
  
  purrr::map_dfr(
    .x = 2011:2020,
    .f = function(current_lau_year) {
      purrr::map_dfr(
        .x = available_pop_grid_years,
        .f = function(current_pop_grid_year) {
          dplyr::semi_join(x = get_pop_grid_countries(current_pop_grid_year, db),
                           y = get_lau_countries(current_lau_year, db),
                           by = "country") %>% 
            dplyr::arrange(country) %>% 
            dplyr::add_count(name = "n") %>% 
            dplyr::summarise(country = stringr::str_c(country, collapse = ", "),
                             n = unique(n)) %>% 
            dplyr::mutate(lau_year = current_lau_year, pop_grid_year = current_pop_grid_year)
        }
      )
    })





pop_grid_lau_countries

#pop_grid_lau_countries <- c("LU", "CY", "MK", "IS", "RS")

pop_grid_lau_countries %>% 
  dplyr::select(lau_year, pop_grid_year, n, country) %>% 
  dplyr::arrange(lau_year, pop_grid_year) %>% 
  knitr::kable()
```

## Availability of LAU/NUTS correspondence tables

See [source for details](https://ec.europa.eu/eurostat/web/nuts/local-administrative-units)

